--- 
title: "Case studies with mrgsolve: PBPK and QSP model implementation and utilization in R"
date: ""
author: ""
output:
  github_document:
    toc: false
---

```{r global, message = FALSE, echo = FALSE, warning = FALSE}
library(mrgsolve)
library(tidyverse)
knitr::opts_chunk$set(fig.path = "docs/img/README-", message = FALSE, warning = FALSE)
```



```{r, echo = FALSE, message = FALSE, warning = FALSE}

mod <- mread("rifampicin_midazolam", "docs/models") %>% param(WT = 90)


rif <- ev(amt = 75, ii = 24, addl = 6, cmt = 1)
mid <- ev(amt = 3, cmt = 2)

comb <- ev_seq(rif, wait = -12, mid)
mono <- filter(as.data.frame(comb), cmt==2) %>% as.ev

out1 <- 
  mrgsim(mod, events = comb, delta = 0.1, end = 240) %>%
  mutate(tx = 2)

out2 <- 
  mrgsim(mod, events = mono, delta = 0.1, end = 240) %>% 
  mutate(tx = 1)

sims <- bind_rows(out1,out2) %>% 
  mutate(tx = factor(
    tx, labels = c("No Rifampicin", "Rifampicin Pretreatment x7d")
  ))


simddi <- filter(sims,time >= 156 & time <= 166) %>%
  mutate(time = time - min(time))

ggplot(simddi, aes(time,Cmidazolam, col = factor(tx))) + 
  geom_line(lwd = 1.25)  + 
  scale_color_brewer(palette="Set2", name = "") +
  xlab("Time after dose (hours)") + 
  theme_bw() + theme(legend.position = "top") +
  scale_x_continuous(breaks=seq(0,10,2)) +
  scale_y_continuous(trans = "log10", 
                     name = "Midazolam concentration (ng/mL)",
                     breaks = 10^seq(-4,4), 
                     limits = c(0.1, 10))

```

# Location

- University of Florida - College of Pharmacy (Orlando Lake Nona Campus) 
- Saturday, 24 March 1:30 pm to 5:30 pm

# Content

Please see the reading list below for references for this content

__QSP Models__

- [ERK inhibition in BRAF(V600E) mutant colorectal cancer](docs/mapk_inhibitors_in_colorectal_cancer.md)
- [OpenBoneMin: BMD changes under denosumab treatment](docs/OpenBoneMin.md)

__PBPK Models__

- [PBPK model for rifampicin / midazolam DDI](docs/rifampin_midazolam_ddi.md)
- [Parameter optimization in PBPK model for statin/CsA DDI](docs/oatp_ddi_optimization.md)

__Methods__

- [Global sensitivity analysis](docs/global_sensitivity_analysis.md)

# Reading List

- __Quantitative Analyses of Hepatic OATP-Mediated Interactions Between 
  Statins and Inhibitors Using PBPK Modeling With a Parameter Optimization 
  Method__
    - Yoshikado T, Yoshida K, Kotani N, Nakada T, Asaumi R, Toshimoto K, Maeda K,
      Kusuhara H, Sugiyama Y. Clin Pharmacol Ther. 2016 Nov;100(5):513-523. 
      PMID: 27170342
    - https://www.ncbi.nlm.nih.gov/pubmed/27170342

- __Comprehensive PBPK Model of Rifampicin for Quantitative Prediction of
Complex Drug-Drug Interactions: CYP3A/2C9 Induction and OATP Inhibition Effects__
    - Asaumi R, Toshimoto K, Tobe Y, Hashizume K, Nunoya KI, Imawaka H, Lee W,
      Sugiyama Y.  CPT Pharmacometrics Syst Pharmacol. 2018 Jan 25. 
      PMID: 29368402
      - https://www.ncbi.nlm.nih.gov/pubmed/29368402

- __Clinical responses to ERK inhibition in BRAF(V600E)-mutant
colorectal cancer predicted using a computational model__
    - Kirouac DC, Schaefer G, Chan J, Merchant M, Orr C, Huang SA, Moffat J, Liu L, 
      Gadkar K, Ramanujan S.  NPJ Syst Biol Appl. 2017 Jun 2;3:14. 
      PubMed PMID: 28649441
    - https://www.ncbi.nlm.nih.gov/pubmed/28649441

- __Predicting nonlinear changes in bone mineral density
over time using a multiscale systems pharmacology model__
    -  Peterson MC, Riggs MM. CPT Pharmacometrics Syst
       Pharmacol. 2012 Nov 14;1:e14. 
       PMID: 23835796
    - https://www.ncbi.nlm.nih.gov/pubmed/23835796

- __Sobol Sensitivity Analysis: A Tool to Guide the Development and Evaluation 
of Systems Pharmacology Models__
    - Zhang XY, Trame MN, Lesko LJ, Schmidt S. . CPT
      Pharmacometrics Syst Pharmacol. 2015 Feb;4(2):69-79.  
      PMID: 27548289
    - https://www.ncbi.nlm.nih.gov/pubmed/27548289
    
    