---
title: Sobol sensitivity analysis using sensobol
author: "Metrum Research Group"
output:
  github_document:
    toc: TRUE
---

```{r, echo = FALSE, message = FALSE}
library(here)

knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  comment = '.', 
  fig.path = here("docs/img/sobol")
)
```


# Reference / About
Zhang XY, Trame MN, Lesko LJ, Schmidt S. __Sobol Sensitivity Analysis: A Tool to
Guide the Development and Evaluation of Systems Pharmacology Models__. CPT
Pharmacometrics Syst Pharmacol. 2015 Feb;4(2):69-79. doi: 10.1002/psp4.6. PubMed
PMID: [27548289](https://www.ncbi.nlm.nih.gov/pubmed/27548289)

This example replicates an analysis presented in the Zhang et al. paper, 
but here using mrgsolve and other tools available for R.

__NOTE__: This example uses the `sensobol` package to run the analysis. This is 
my preferred package for global sensitivity analysis. You can 
see the same analysis run with `sensitivity` here: [sobol.md](sobol.md). 

# Introduction 

This vignette shows you how to to Sobol sensitivity analysis, a form of global
sensitivity analysis. In this type of analysis, we look at the relationship 
between variability in input parameters and variability in model outputs when 
model parameters are all varied together across the plausible parameter space. 
This is in contrast to local sensitivity analysis, where parameters are 
perturbed one at a time by small amounts. 

To do Sobol sensitivity analysis, we will need the following

1. A model with some parameters to manipulate 
1. A (large) set of random parameter values covering a certain section of the 
plausible (or reasonable) parameter space
1. An intervention of interest (e.g. a dose)
1. A model output, summarized as a single value, that depends on the 
intervention and the value of the manipulated parameters (e.g. AUC for a 
PK model)
1. Calculate sensitivity indices, which summarize the relationships between 
input parameters and model outputs
1. Visualization or presentation of the indices


# Tools
```{r setup}
library(mrgsolve)
library(tidyverse)
library(data.table)
library(sensobol)
library(mrgmisc)
```

Install `mrgmisc` 

```{r, eval = FALSE}
install.packages(
  "mrgmisc", 
  repos = "https://mpn.metworx.com/snapshots/stable/2023-10-19"
)
```


# The sunitinib PK model
```{r}
mod <- mread("sunit", here("docs/models")) %>% 
  update(end = 24, delta = 0.5, outvars = "CP") %>% 
  zero_re()
```

```{r}
see(mod)
```

## Sunitinib dosing 

We are just looking at a single dose for now.

```{r}
sunev <- function(amt = 50,...) ev(amt = amt, ...)
```

# Generate samples

__Number__: sample size of the base sample matrix

```{r}
N <- 2 ^ 15
```

__Generate__

Generate the parameter sets. 

```{r}
mat <- sobol_matrices(N = N, params = c("TVCL", "TVVC", "TVKA", "TVQ", "TVVP"))
head(mat)
```

Take a look at the package vignette for `sensobol` for better details. Briefly, 
this is a matrix with random variates for each parameter in the sensitivity
analysis. The samples are uniform between zero and one. You will have to
transform these variates to parameter values as we'll see in the next section. 

__Transform and groom__

For this example, we will assume that all parameters have uniform distribution
between 1/5 and 5 times the current value of the parameter

```{r}
params <- unlist(as.list(param(mod))[c("TVCL", "TVVC", "TVKA", "TVQ", "TVVP")])
umin <- params / 5
umax <- params * 5
umin
umax
```

After setting the minimum and maximum for each parameter, get the value of the 
parameter by looking at the quantiles of the uniform distribution (`qunif()`)

```{r}
mat <- as_tibble(mat)
mat2 <- imodify(mat, ~ qunif(.x, umin[[.y]], umax[[.y]]))
mat2 <- mutate(mat2, ID = row_number())
head(mat2)
```


If we wanted each parameter to have log-normal distribution with 30% coefficient
of variation, we pass our uniform (0,1) variates into `qlnorm()` instead

```{r}
mat3 <- imodify(mat, ~ qlnorm(.x, log(params[[.y]]), sqrt(0.09)))
summary(mat3)
params
```

# Run the analysis
We will use the uniform parameters that we generated above.

## Simulation
Now, we have a set of parameters for the model and we can simulate PK over
24 hours using `mrgsim_ei()`, passing in an event object and the idata set

```{r}
out <- mrgsim_ei(mod, sunev(), mat2, output = "df")
out <- as.data.table(out)
```

## Calculate AUC
```{r}
y <- out[, list(auc = auc_partial(time, CP)), by = "ID"][["auc"]]
```

## Indices
```{r}
ind <- sobol_indices(Y = y, N = N, params = names(params), boot = TRUE, R = 1000, first = "jansen")
ind.dummy <- sobol_dummy(Y = y, N = N, params = names(params), boot = TRUE, R = 1000)
```

```{r}
ind
```


## Visualize
First, plot uncertainty in outcome

```{r}
plot_uncertainty(Y = y, N = N) + scale_x_log10()
```

Now, plot indices

```{r}
plot(ind, dummy = ind.dummy) + ylim(0,1)
```

Plot outputs versus inputs

```{r}
plot_scatter(N = 2000, data = mat2, Y = y, params = names(mat))
```

```{r, fig.height = 8}
plot_multiscatter(N = 2000, data = mat2, Y = y, params = names(mat))
```
