---
title: Sobol sensitivity analysis
author: "Metrum Research Group, LLC"
date: ""
output:
  github_document:
    toc: TRUE
---

```{r, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, 
                      comment = '.', fig.path = "img/sobol")
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.path = "cache/sobol/")
```


# Reference / About
Zhang XY, Trame MN, Lesko LJ, Schmidt S. __Sobol Sensitivity Analysis: A Tool to
Guide the Development and Evaluation of Systems Pharmacology Models__. CPT
Pharmacometrics Syst Pharmacol. 2015 Feb;4(2):69-79. doi: 10.1002/psp4.6. PubMed 
PMID: [27548289](https://www.ncbi.nlm.nih.gov/pubmed/27548289)

This example replicates an analysis presented in the Zhang et al. paper, 
but here using mrgsolve and other tools available for R.


# Tools
```{r}
library(mrgsolve)
library(tidyverse)
library(PKPDmisc)
library(sensitivity)
```



# The sunitinib PK model
```{r}
mod <- mread_cache("sunit", "models") %>% 
  update(end = 24, delta = 1) %>% zero_re()
```

```{r}
see(mod)
```

## Sunitinib dosing 
```{r}
sunev <- function(amt = 50,...) ev(amt = amt, ...)
```

## Generate samples

Th function generates uniform samples from a 
100 fold decrease to 100 fold increase in the 
nominal parameter value.

The return value is a list with two data frames that
can be passed into the sobol function.


```{r}
gen_samples <- function(n, l, which = names(l), 
                        factor = c(0.01,100)) {
  
  vars <- select_vars(names(l), !!(enquo(which)))
  
  l <- as.list(l)[vars]
  
  l <- map(l, .f = function(x) x*factor)

  n <- length(l)*n*2
  
  df <- as.data.frame(l)
  
  len <- length(df)
  
  X <- matrix(ncol=len, nrow=n)
  
  colnames(X) <- names(df)
  
  Y <- X
  
  for(i in seq(len)){
    r <- runif(n, df[1,i], df[2,i])
    X[,i] <- r
    r <- runif(n, df[1,i], df[2,i])
    Y[,i] <- r
  }
  
  return(list(x1 = as.data.frame(X), x2 = as.data.frame(Y)))
}
```

## A bunch of helper functions

Simulate a batch of data.  The summary is AUC for each parameter set.

```{r}
batch_run <- function(x) {
  mod %>% 
    idata_set(x) %>%
    ev(sunev()) %>%
    mrgsim(obsonly = TRUE) %>% 
    group_by(ID) %>% 
    summarise(AUC = auc_partial(time,CP)) %>% 
    pull(AUC)
}
```



## Run the analysis

### First, generate the samples
```{r}
set.seed(88771)
samp <- gen_samples(6000, param(mod), TVCL:TVVP)

head(samp$x1)
dim(samp$x1)
```


### Then, run `sensitivity::sobol2007`
```{r}
x <- sobol2007(batch_run, X1=samp$x1, X2=samp$x2, nboot=100)
```

### Results

```{r}
plot(x)
```

```{r}
x
```

# The HIV model


```{r}
mod <- mread_cache("hiv", "models") %>% 
  update(end = 2000, delta = 1000, maxsteps = 50000)


out <- mrgsim(mod, 
              idata = data_frame(N = c(1000,1200,1400)),
              end = 10*365, delta = 0.1) 

plot(out, V+L+I+TAR~time/365)


```


```{r}
bound <- tribble(
~name , ~lower   , ~upper,
"s"	  , 1.00E-02 , 50,
"muT" ,	1.00E-04 , 0.2,
"r"	  , 1.00E-03 , 50,
"k1"  ,	1.00E-07 , 1.00E-03,
"k2"  , 1.00E-05 , 1.00E-02,
"mub" , 1.00E-01 , 0.4,
"N"   , 1        , 2000,
"muV" ,	1.00E-01 , 10
)

mksamp <- function(bounds, n) {
  x <- split(bounds,seq(nrow(bounds)))
  out <- map(x, .f = function(xx) {
    runif(n, xx$lower[1], xx$upper[1])  
  })
  names(out) <- bounds$name
  return(as_data_frame(out))
}

set.seed(10010)
x1 <- as.data.frame(mksamp(bound,4000*nrow(bound)))
x2 <- as.data.frame(mksamp(bound,4000*nrow(bound)))
```




```{r}
hiv_run <- function(x) {
  
  out <- mrgsim_i(x = mod, idata = x)
  
  out %>% filter(time==2000) %>% pull(AUC)
}
```

```{r}
x <- sobol2007(hiv_run, X1=x1, X2=x2, nboot=100)
```


```{r}
tot <- x$T %>% mutate(type = "total order",   parameter = names(x1))

first <- x$S %>% mutate(type = "first order", parameter = names(x1))

sum <- bind_rows(tot,first) %>% mutate(ymax = original + 1.96*`std. error`)

ggplot(data = sum, aes(x = parameter, y = original, fill = type)) + 
  geom_col(position = position_dodge()) + 
  geom_errorbar(aes(ymin = original, ymax = ymax), position = position_dodge()) + 
  scale_fill_brewer(palette = "Set2", name = "") + 
  theme_bw() + ylab("Sensitivity indices") +
  theme(legend.position = "top") +
  scale_y_continuous(limits = c(0,1), breaks = seq(0,1,0.1))

```



# Session
```{r}
devtools::session_info()
```


