---
title: "Clinical responses to ERK inhibition in BRAF{V600E}-mutant colorectal cancer"
author: "Metrum Research Group, LLC"
date: ""
output: 
  github_document:
    toc: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(comment = '.', fig.path = "img/mapk-", 
                      message = FALSE, warning = FALSE)

```


```{r}
library(mrgsolve)
library(tidyverse)
source("functions.R")
```

__Clinical responses to ERK inhibition in BRAF{V600E}-mutant colorectal cancer predicted using a computational model__

- Daniel C. Kirouac, Gabriele Schaefer, Jocelyn Chan, Mark Merchant, Christine
Orr, Shih-Min A. Huang, John Moffat, Lichuan Liu, Kapil Gadkar and Saroja
Ramanujan

- npj Systems Biology and Applications  (2017) 3:14 ; doi:10.1038 / 
s41540-017-0016-1

# About

(Summarized from Introduction in the reference)

- The V600E/K mutation results in constitutively 
active BRAF, with subsequent signalling through MEK
and ERK

- BRAF and MEK inhibitors were found to be effective in 
V600E mutant melanoma, but not so much in colorectal cancer
    - Could resistance to BRAF inhibitors be mediated through 
      EGFR signalling through RAS and CRAF?
    - What about inhibition at ERK?

- Could the effectiveness of different combination therapies be predicted
with a model characterizing this biology?

## Cast of characters
- __vemurafenib__: BFAF inhibitor (selective for V600E mutant)
- __cobimetinib__: MEK inhibitor
- __cetuximab__: EGFR antibody
- __GDC-0994__: ERK inhibitor (the star)

# Translation 

- Model published as SBML
- Translator from previous project work using R bindings to libSBML
- Minor modifications to the translator code to
  accommodate the MAPK model as published

# Explore

Read in the virtual population

```{r}
vp <- readRDS("data/s10vpop_pk.RDS") %>% mutate(VPOP2 = seq(n()))

dim(vp)
```

Load the model and pick one parameter set from vpop
```{r, message = FALSE}
mod <- mread_cache("mapk", "models") %>% update(end  = 56)

mod <- param(mod, filter(vp,VPOP2==24))
```

```{r}
e <- expand.ev(amt = c(1,3,10,30,60,100), cmt = 10, ii = 1, addl = 20)

e <- ev_seq(e, wait = 7, e) %>% as_data_frame %>%  arrange(ID)

mod %>% 
  data_set(e) %>%
  Req(TUMOR,MEKi) %>%
  mrgsim(delta = 0.25) %>% 
  plot(MEKi+TUMOR ~ time)
```

- Look at MAPK dependence
```{r}
library(mrgsolvetk)

dat <- filter(e, amt==60) %>% select(-ID) %>% as.ev

mod %>% 
  ev(as.ev(dat)) %>% Req(TUMOR) %>%
  sens_spaced(wOR = c(0.9,0.99), .n = 10) %>%
  ggplot(aes(time, TUMOR, col = factor(value), group = value)) + 
  geom_line(lwd = 1) + geom_hline(yintercept = 1, lty = 2) + theme_bw()
  
```

- What about $\delta_{max}$ dependence?


```{r}
mod %>% 
  ev(as.ev(dat)) %>% Req(TUMOR) %>%
  select(dmax) %>% 
  sens_spaced_factor(.factor = 2, .n = 10) %>%
  mutate(value = signif(value,3)) %>%
  ggplot(aes(time, TUMOR, col = factor(value), group = value)) + 
  geom_line(lwd = 1) + geom_hline(yintercept = 1, lty = 2) + theme_bw()

```


# Combination therapies

- Re-create figure 6B in the publication 



Update the model object with one set of parameters

```{r}
mod <- param(mod, slice(vp, 1))
```

```{r}
summary(vp$wOR)
summary(vp$dmax)
```


## Generate dosing regimens


- __No treatment__

```{r}
data0 <- ev(amt=0, cmt=8)
```

- __BRAF inhibitor__ - vemurafanib (VEMU)
- Compartment 8
```{r}
dataV <- ev(amt=960,  cmt=8, ii=0.5, addl=120)
```

- __ERK inhibitor__ - GCD-994 (GDC)
- Compartment 12
```{r}
dataG <- ev(amt = 400, cmt = 12, ii = 1, addl = 20)

dataG <- seq(dataG, wait = 7, dataG) 

out <- mrgsim(mod, ev=dataG, end=56)

plot(out, ERKi_C~time)
```


- __MEK inhibitor__ - cobimetinib (COBI)
- Compartment 10
```{r}
dataCO <- mutate(dataG,amt=60,cmt=10)
```

- __EGFR inihbitor__ - cetuximab (CETUX)
- Compartment 7
```{r}
dataCE <- ev(cmt=7,ii=7,addl=7,amt=450)
```




We create two functions: one to combine dosing regimens
and the other to simulate from a dosing regimen
```{r}
comb <- function(...) {
  x <- lapply(list(...), as.data.frame)
  bind_rows(x) %>% arrange(time)
}

sim <- function(Data,Vp,Mod) {
  Mod %>%
    ev(as.ev(Data)) %>%
    mrgsim(idata=Vp,end=-1, add = 56) %>%
    filter(time==56) 
}
```

For example

```{r}
comb(dataCE,dataV)

sim(comb(dataCE,dataV), Vp = slice(vp,seq(10)), Mod = mod)
```



Generate a data frame of runs to do

```{r}
sims <- 
  tribble(
    ~label, ~object, 
    "No Treatment",        data0,
    "CETUX",               dataCE, 
    "VEMU",                dataV,
    "COBI",                dataCO, 
    "GDC",                 dataG,
    "CETUX+VEMU",          comb(dataCE, dataV), 
    "CETUX+COBI",          comb(dataCE, dataCO), 
    "CETUX+GDC",           comb(dataCE, dataG),
    "VEMU+COBI",           comb(dataV, dataG), 
    "VEMU+GDC",            comb(dataV, dataG),
    "COBI+GDC",            comb(dataCO, dataG),
    "CETUX+VEMU+COBI",     comb(dataCE, dataV,dataCO), 
    "CETUX+VEMU+GDC",      comb(dataCE, dataV,dataG), 
    "CETUX+COBI+GDC",      comb(dataCE, dataCO,dataG), 
    "VEMU+COBI+GDC",       comb(dataV, dataCO,dataG),
    "CETUX+VEMU+COBI+GDC", comb(dataCE, dataV, dataCO, dataG)
  ) %>% mutate(object = map(object,as.data.frame))

```


Run the simulation

```{r}
sims <- mutate(sims, out = parallel::mclapply(object,sim, Vp = vp, Mod = mod))
```


Get ready to plot
```{r}
sms <- select(sims, label, out) %>% unnest()

sms <- mutate(sms, labelf = fct_inorder(label))

head(sms)
```


```{r}
p1 <- 
  ggplot(data=sms) + 
  geom_point(aes(x=labelf, y=TUMOR),position=position_jitter(width=0.15),col="grey") +
  scale_y_continuous(limits=c(0,2.5),name="Tumor size",breaks=c(0,0.5,1,1.5,2,2.5,3)) +
  scale_x_discrete(name="") + 
  geom_hline(yintercept=0.7,col="firebrick", lty=1,lwd=1)  +
  geom_boxplot(aes(x=labelf,y=TUMOR),fill="darkslateblue",col="darkslateblue",alpha=0.2) +
  theme_plain() + rotx(30)
p1
```

