---
title: "Clinical responses to ERK inhibition in BRAF{V600E}-mutant colorectal cancer"
author: "Metrum Research Group"
date: ""
output: 
  github_document:
    toc: true
---

```{r, echo = FALSE, message = FALSE}
library(here)

knitr::opts_chunk$set(
  comment = '.', 
  fig.path = here("docs/img/mapk-"), 
  cache.path = here("docs/cache/mapk/"),
  message = FALSE, 
  warning = FALSE,
  cache = TRUE, 
  autodep = TRUE 
)

options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(name = "Dark2", n = 8))
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(name = "Dark2", n = 8))
```


# Reference 

__Clinical responses to ERK inhibition in BRAF{V600E}-mutant colorectal cancer 
predicted using a computational model__

- Daniel C. Kirouac, Gabriele Schaefer, Jocelyn Chan, Mark Merchant, Christine
Orr, Shih-Min A. Huang, John Moffat, Lichuan Liu, Kapil Gadkar and Saroja
Ramanujan

- npj Systems Biology and Applications (2017) 3:14 ; doi:10.1038 / 
s41540-017-0016-1

# Introduction

(Summarized from Introduction in the reference)

- The V600E/K mutation results in constitutively 
active BRAF, with subsequent signalling through MEK
and ERK

- BRAF and MEK inhibitors were found to be effective in 
V600E mutant melanoma, but not so much in colorectal cancer
- Could resistance to BRAF inhibitors be mediated through 
EGFR signalling through RAS and CRAF?
- What about inhibition at ERK?
- Could the effectiveness of different combination therapies be predicted
with a model characterizing this biology?

## Cast of characters
- __vemurafenib__: BRAF inhibitor (selective for V600E mutant)
- __cobimetinib__: MEK inhibitor
- __cetuximab__: EGFR antibody
- __GDC-0994__: ERK inhibitor (the star)

## Translation 

- Model published as SBML
- Translator from previous project work using R bindings to libSBML
- Minor modifications to the translator code to accommodate the MAPK model as 
published

# Set up
```{r}
library(mrgsolve)
library(tidyverse)
library(parallel)
library(mrgsim.sa)
library(future.apply)
library(knitr)
source(here("docs/functions.R"))
```

Read in the virtual population

```{r}
vp <- readRDS(here("docs/data/s10vpop_pk.RDS")) %>% mutate(VPOP2 = seq(n()))

dim(vp)

```

Load the model and pick one parameter set from vpop

```{r, message = FALSE}
mod <- mread("mapk", here("docs/models"), soloc = here("docs/build"), end = 56) 

mod <- param(mod, filter(vp, VPOP2==41))
```


# Explore


## Simulate with ERK inhibitor GDC-0944

```{r, fig.width = 8}
e <- expand.ev(amt = seq(100,600,100), cmt = 12, ii = 1, addl = 20) %>% as.ev()

e <- ev_seq(e, wait = 7, e) %>% as_tibble() %>% arrange(ID)

mod %>% 
  data_set(e) %>%
  mrgsim(delta = 0.25) %>% 
  plot(ERKi + TUMOR ~ time)
```

## Sensitivity analysis


The authors note two parameters that are particularly influential with respect 
to response rates:

- `wOR`: MAPK pathway dependence parameter
- $\delta_{max}$: the maximum cell death rate


```{r}
vp %>% 
  select(wOR,dmax) %>%
  map(quantile, probs = seq(0,1,0.1)) %>% 
  bind_cols() %>% 
  mutate(pctile = seq(0,1,0.1))
```


__Sensitivity analysis on MAPK pathway dependence__

- Given the other parameters for this virtual patient, whenever 
`wOR` falls below 90% or so, you get no change in tumor size

- The authors discuss the possibility of looking for biomarkers that would 
indicate a patient is showing high dependence on MAPK signalling to improve
response rates

```{r}
#| warning: false
ev400 <- filter(e, amt==400) %>% select(-ID) %>% as.ev()

mod %>% 
  ev(ev400) %>% 
  Req(TUMOR) %>%
  parseq_range(wOR = c(0.9,1.05), .n = 8) %>%
  sens_each() %>% 
  sens_plot("TUMOR")
```


__Sensitivity analysis on $\delta_{max}$__

- In this analysis, we look between 0.2 and 2 times the nominal 
parameter value for this virtual patient

- The authors do a simulation with a hypothetical agent that increases the 
cell death rate by 10%

```{r}
#| warning: false
mod %>% 
  ev(ev400) %>% 
  Req(TUMOR) %>%
  parseq_fct(dmax, .n = 8) %>%
  sens_each() %>% 
  sens_plot("TUMOR")
```


# Predicting clinical outcomes for combination therapies

- Re-create figure 6B in the publication 


## Generate dosing regimens


- __No treatment__

```{r}
data0 <- ev(amt = 0, cmt = 8)
```

- __BRAF inhibitor__ - vemurafanib (VEMU)
- Compartment 8
```{r}
dataV <- ev(amt = 960, cmt = 8, ii = 0.5, addl = 120)
```

- __ERK inhibitor__ - GCD-994 (GDC)
- Compartment 12
```{r}
dataG <- ev(amt = 400, cmt = 12, ii = 1, addl = 20)

dataG <- seq(dataG, wait = 7, dataG) 

out <- mrgsim(mod, ev = dataG, end = 56, delta = 0.1)

plot(out, ERKi_C~time)
```


- __MEK inhibitor__ - cobimetinib (COBI)
- Compartment 10
```{r}
dataCO <- mutate(dataG, amt = 60, cmt = 10)
```

- __EGFR inihbitor__ - cetuximab (CETUX)
- Compartment 7
```{r}
dataCE <- ev(cmt = 7, ii = 7, addl = 7, amt = 450)
```




We create two functions: one to combine dosing regimens
and the other to simulate from a dosing regimen
```{r}
comb <- function(...) {
  x <- lapply(list(...), as.data.frame)
  bind_rows(x) %>% arrange(time)
}

sim <- function(d, m, v) {
  m %>%
    ev(as.ev(d)) %>%
    mrgsim(idata = v, end = -1, add = 56) %>%
    filter(time==56) 
}
```

For example, to dose cetuximab with vemurafanib, we'd do

```{r}
comb(dataCE, dataV)

sim(d = comb(dataCE,dataV), m = mod, v = slice(vp, seq(10)))
```


## Simulate all combination therapies

Generate a data frame of runs to do

```{r}
sims <- 
  tribble(
    ~label, ~object, 
    "No Treatment",        data0,
    "CETUX",               dataCE, 
    "VEMU",                dataV,
    "COBI",                dataCO, 
    "GDC",                 dataG,
    "CETUX+VEMU",          comb(dataCE, dataV), 
    "CETUX+COBI",          comb(dataCE, dataCO), 
    "CETUX+GDC",           comb(dataCE, dataG),
    "VEMU+COBI",           comb(dataV, dataG), 
    "VEMU+GDC",            comb(dataV, dataG),
    "COBI+GDC",            comb(dataCO, dataG),
    "CETUX+VEMU+COBI",     comb(dataCE, dataV,dataCO), 
    "CETUX+VEMU+GDC",      comb(dataCE, dataV,dataG), 
    "CETUX+COBI+GDC",      comb(dataCE, dataCO,dataG), 
    "VEMU+COBI+GDC",       comb(dataV, dataCO,dataG),
    "CETUX+VEMU+COBI+GDC", comb(dataCE, dataV, dataCO, dataG)
  ) %>% mutate(object = map(object, as.data.frame))

```


Run the simulation

```{r simulate}
plan(multisession, workers = 5L)

sims <- mutate(
  sims, 
  out = future_lapply(X = object, FUN = sim, v = vp, m = mod, future.seed = TRUE),
)
```

## Summarize and plot

Get ready to plot
```{r}
sms <- select(sims, label, out) %>% unnest(cols = c(out))

sms <- mutate(
  sms, 
  labelf = fct_inorder(label), 
  gdc = factor(grepl("GDC", label))
)

head(sms)
```


```{r, fig.width = 9}
p1 <- 
  ggplot(data=sms) + 
  geom_point(aes(x=labelf, y=TUMOR),position=position_jitter(width=0.15),col="grey") +
  geom_hline(yintercept=0.7,col="black", lty=1,lwd=0.7)  +
  scale_y_continuous(limits=c(0,2.5),name="Tumor size",breaks=c(0,0.5,1,1.5,2,2.5,3)) +
  scale_x_discrete(name="") + 
  geom_boxplot(aes(x=labelf,y=TUMOR,col=gdc),fill="darkslateblue",alpha=0.2) +
  scale_color_manual(values = c("darkslateblue", "firebrick"), guide = "none") + 
  theme_plain() + rotx(30) + 
  ggtitle("Note: GDC-0944 +/- cobimetinib")

p1
```

# Target populations more likely to respond

## ORR in full population: GDC +/- COBI

```{r}
sms %>%
  filter(label %in% c("GDC", "COBI+GDC")) %>%
  group_by(label) %>%
  summarise(orr = mean(TUMOR < 0.7)) %>% 
  kable(digits = 3)
```

## ORR in select patients: GDC +/- COBI

- We filter down to those with `wOR` greater than the median

```{r}
vp_select <- filter(vp, wOR > median(wOR))
```

And resimulate

```{r resimulate with select population}
plan(sequential)
plan(multisession, workers = 5L)

re_run <- 
  sims %>%
  select(label, object) %>%
  filter(label %in% c("GDC", "COBI+GDC")) %>% 
  mutate(
    out = future_lapply(object, sim, v = vp_select, m = mod, future.seed = TRUE)
  ) %>%
  select(label, out) %>% 
  unnest(cols = c(out))
```

Now, response rates are up

```{r}
re_run %>%
  group_by(label) %>%
  summarise(orr = mean(TUMOR < 0.7)) %>% 
  kable(digits = 3)
```

