---
title: "Fit parameters in PBPK model"
author: "Metrum Research Group"
date: ""
output: 
  github_document:
    toc: true
    toc_depth: 2
---

# Packages and setup

```{r, message = FALSE}
library(tidyverse)
library(mrgsolve)
library(pmxTools)
library(minqa)
library(RcppDE)
library(GenSA)
library(hydroPSO)
library(here)
library(knitr)

source(here("docs/functions.R"))

set.seed(10101)

theme_set(theme_bw() + theme(legend.position="top"))

options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(name = "Dark2", n = 8))
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(name = "Dark2", n = 8))
```

```{r, echo = FALSE}
knitr::opts_chunk$set(comment = '.', message = FALSE, warning = FALSE) 
knitr::opts_chunk$set(
  fig.path = "img/oatp_ddi_optimization-"
) 
knitr::opts_chunk$set(
  cache = TRUE, 
  autodep = TRUE, 
  cache.path = "cache/optim/"
)
```

# Reference
__Quantitative Analyses of Hepatic OATP-Mediated Interactions Between Statins and Inhibitors Using PBPK Modeling With a Parameter Optimization Method__

- T Yoshikado, K Yoshida, N Kotani, T Nakada, R Asaumi, K Toshimoto, K Maeda, 
  H Kusuhara and Y Sugiyama

- CLINICAL PHARMACOLOGY & THERAPEUTICS | VOLUME 100 NUMBER 5 | NOVEMBER 2016 

- https://www.ncbi.nlm.nih.gov/pubmed/27170342


# Data

- Example taken from figure 4a from the publication
- Using this as example data to fit

```{r, message = FALSE}
data <- read_csv(here("docs/data/fig4a-fit.csv")) 

data <- 
  mutate(
    data, 
    type = ID, 
    typef = factor(ID, labels = c("Statin", "Statin+CsA"))
  )

head(data)
```

**Important**: the data has `DV` set to `NA` (missing value); this is important 
to for the following approach to work.


- The goal is to fit the pitavastatin data either alone (left) or in combination 
with cyclosporin administered 1 hour before the pitavastatin

```{r}
ggplot(data = data, aes(TIME, DV)) + 
  geom_point(aes(col = typef), size = 3) + 
  geom_line(col = "darkgrey", aes(group = typef)) + 
  scale_y_continuous(trans = "log", limits = c(0.1,300), breaks = logbr()) 
```


# PBPK model: pitavastatin / CsA DDI

- Check out the model / data with a quick simulation
- We're only interested in `CP`, the pitavastatin concentration

```{r, message = FALSE, cache = FALSE}
mod <- mread_cache("yoshikado", here("docs/models")) 

mod <- update(mod, end = 14, delta = 0.1, outvars = "CP")
```

This model has `r length(param(mod))` parameters and `r length(init(mod))`
compartments

```{r}
param(mod)

init(mod)
```

Simulate out a quick example ... we filter out just the dosing records and 
fill in with a smooth set of observations

```{r}
doses <- filter(data, EVID==1)

sims <- 
  mod %>% 
  mrgsim(data = doses, obsaug = TRUE) %>% 
  mutate(type = typef(ID))

ggplot(sims, aes(TIME, CP, col = type)) + 
  geom_line(lwd = 1) + 
  scale_x_continuous(breaks = seq(0, 12, 2)) + 
  scale_y_continuous(trans = "log10") + 
  ylab("Pitavastatin concentration") + xlab("Time (hour)") 
```


```{r}
sims %>% 
  group_by(type) %>% 
  pmxTools::get_auc(dv = "CP") %>% 
  mutate(fold_increase = AUC/first(AUC))
```


# Objective function

- Least squares objective function
- Weighted by the observations

```{r}
wss <- function(dv, pred, par = NULL) {
  sum(((dv-pred)/dv)^2, na.rm = TRUE)
}
```

### Prediction function

- Let's go through step by step what each line is doing for us

```{r}
pred <- function(p, m, d, pred = FALSE) {
  
  p <- lapply(p, exp)
  
  names(p) <- names(theta)
  
  m <- param(m, p)
  
  if(pred) {
    out <- mrgsim_df(m, data = d, recover = "type")
    return(out)
  }
  
  out <- mrgsim_q(m, data = d, output = "df")
  
  return(wss(d$DV, out$CP))
  
  #-1*sum(dnorm(log(yobs),log(out$CP),.par$sigma,log=TRUE), na.rm=TRUE)
}
```


# Optimize with different methods

## `minqa::newuoa`: minimization without derivatives

- These appear to be the parameters that the authors are fitting

```{r optimize newuoa}
theta <- log(c(fbCLintall = 1.2, ikiu = 1.2, fbile = 0.9, ka = 0.1, ktr = 0.1))

control <- list(iprint = 25)

fit1 <- newuoa(theta, pred, m = mod, d = data, control = control)

fit1$par <- setNames(fit1$par, names(theta))
```

### Get some predictions to look at how the fit went

- Predictions with the final estimates
- Predictions with the initial estimates
- Observed data to overlay

```{r}
df_pred <- pred(fit1$par, mod, doses, pred = TRUE) %>% mutate(type = typef(type))
df_init <- pred(theta, mod, doses, pred = TRUE) %>% mutate(type = typef(type))
df_obs <- mutate(data, type = typef(type))
```

### Plot
```{r}
ggplot(data = df_pred, lwd = 0.7) + 
  geom_line(data = df_init,aes(TIME, CP, lty = "A")) +
  geom_line(aes(TIME, CP, lty = "B")) + 
  geom_point(data = df_obs, aes(TIME, DV, col = type), size = 3) + 
  facet_wrap(~type) + 
  scale_y_log10(limits=c(0.1, 100)) +
  ylab("Pitavastatin concentration (ng/mL)") +
  scale_x_continuous(breaks = seq(0, 14, 2), name = "Time (hours)") +
  scale_linetype_manual(
    values = c(2,1), 
    labels = c("Initial estimates", "Final estimates"), 
    guide = "none",
    name = ""
  )
```

### The final objective function value and estimates

```{r}
pred(fit1$par, m = mod, d = data)

exp(fit1$par)
```

## `optim`: Nelder-Mead

This optimizer comes from the `stats` package in base R. 

```{r optimize nelder mead}
fit1b <- stats::optim(theta, pred, m = mod, d = data)
```

## `DEoptim`: differential evolution algorithm

"Performs evolutionary global optimization via the Differential Evolution 
algorithm."

```{r optimize de optim}
lower <- rep(-6, length(theta)) %>% setNames(names(theta))
upper <- rep( 4, length(theta)) %>% setNames(names(theta))

set.seed(330303)

decontrol <- DEoptim.control(
  NP = 10*length(theta), 
  CR = 0.925, 
  F = 0.85,
  itermax = 90, 
  storepopfrom = 0, 
  trace = 10
)

fit2 <- DEoptim(
  fn = pred, 
  lower = lower, 
  upper = upper, 
  control = decontrol,
  m = mod, 
  d = data
)
```

### Plot the history

We can plot each population from the DE optimization over time as well as 
the best member for each iteration.

```{r, eval = TRUE}
hx <- lapply(fit2$member$storepop, as.data.frame) %>% bind_rows()
hx <- mutate(hx, iteration = rep(1:decontrol$itermax, each = decontrol$NP))
hx <- mutate(hx, pop = rep(1:decontrol$NP, time = decontrol$itermax))
hxm <- pivot_longer(hx, fbCLintall:ktr) %>% mutate(value = exp(value))
best <- 
  fit2$member$bestmemit %>% 
  as_tibble() %>% 
  mutate(iteration = seq(decontrol$itermax))
bestm <- pivot_longer(best, fbCLintall:ktr) %>% mutate(value = exp(value))
```

```{r, eval = TRUE, fig.width = 8}
ggplot(data = hxm) + 
  geom_line(aes(iteration, value, group = pop), col = "darkslateblue") + 
  geom_line(data = bestm, aes(iteration ,value), col = "orange", lwd = 1) + 
  scale_y_log10(name = "Parameter value") + 
  facet_wrap(~name, ncol = 2, scales = "free_y")
```


## `GenSA`: simulated annealing 

```{r optimize simulated annealing}
set.seed(11001)

sacontrol <- list(maxit = 100, nb.stop.improvement = 20, verbose = TRUE)

fit3 <- GenSA(NULL, pred, lower, upper, m = mod, d = data, control = sacontrol)
```

## `hydroPSO`: particle swarm optimization
```{r optimize particle swarm}
set.seed(2202201)

fit4 <- hydroPSO(
  theta, 
  fn = pred,
  lower = lower, 
  upper = upper, 
  control = list(maxit = 100, REPORT = 5),
  m = mod, d = data
)
```


# Compare optimization methods
```{r}
results <- list(theta, fit1$par, fit1b$par, fit2$optim$bestmem, fit3$par, fit4$par)

results <- map(results, exp)

tibble(
  method = c("initial", "newuoa", "nelder", "RcppDE", "SA", "PSO"),
  fbCLintall = map_dbl(results, "fbCLintall"), 
  ikiu = map_dbl(results, "ikiu"), 
  fbile = map_dbl(results, "fbile"), 
  ka = map_dbl(results, "ka"), 
  ktr = map_dbl(results, "ktr")
) %>% kable(digits = 4)
```
