---
title: "Fit parameters in PBPK model"
author: "Metrum Research Group, LLC"
date: ""
output: 
  github_document:
    toc: true
    toc_depth: 2
---

# Packages and setup

```{r, message = FALSE}
library(tidyverse)
library(mrgsolve)
library(minqa)
library(RcppDE)
library(GenSA)
library(hydroPSO)
source("functions.R")
set.seed(10101)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(comment = '.', message = FALSE, warning = FALSE) 
knitr::opts_chunk$set(fig.path = "img/oatp_ddi_optimization-") 
```

# Reference
__Quantitative Analyses of Hepatic OATP-Mediated Interactions Between Statins and Inhibitors Using PBPK Modeling With a Parameter Optimization Method__

- T Yoshikado, K Yoshida, N Kotani, T Nakada, R Asaumi, K Toshimoto, K Maeda, H Kusuhara and Y Sugiyama

- CLINICAL PHARMACOLOGY & THERAPEUTICS | VOLUME 100 NUMBER 5 | NOVEMBER 2016 

- https://www.ncbi.nlm.nih.gov/pubmed/27170342


# Data

- Example taken from figure 4a from the publication
- Using this as example data to fit
```{r, message = FALSE}
data.file <- file.path("data", "fig4a.csv")

data <- read_csv(data.file) %>% 
  mutate(profile = NULL, 
         type=ID, 
         typef=factor(ID, labels = c("Statin", "Statin+CsA")))
```

- The goal is to fit the pitavastatin data either alone (left) or 
in combination with cyclosporin administered 1 hour before the 
pitavastatin

```{r}
ggplot(data=data,aes(time,DV)) + 
  geom_point(col="firebrick") + 
  facet_wrap(~typef) + 
  scale_y_continuous(trans="log", limits=c(0.1,300), breaks=logbr()) + 
  theme_bw()
```

- Pick out the doses and the observations

```{r}
dose <- filter(data,evid==1) %>% mutate(typef=NULL)

dose 

yobs <- filter(data,evid==0) %>% dplyr::select(DV) %>% unlist %>% unname

yobs

data <-  dplyr::select(data, -typef)
```



# PBPK model: pitavastatin / CsA DDI

- Check out the model / data with a quick simulation
- We're only interested in `CP`, the pitavastatin concentration

```{r, message = FALSE}
mod <- mread("yoshikado","models") %>% 
  update(end=14, delta=0.1) %>% Req(CP) %>% obsonly()

mod %>% 
  mrgsim(data=dose,obsaug=TRUE) %>% 
  plot(CP~.,scales=list(y=list(log=TRUE)))
```

# Objective function

- Least squares objective function
- Weighted by the observations

```{r}
wss <- function(dv, pred, par=NULL) {
  sum(((dv-pred)/dv)^2)
}
```

### Prediction function

- Let's go through step by step what each line is doing for us

```{r}
pred <- function(p, .data, yobs=NULL, pred=FALSE) {
  
  .par <- lapply(p,exp) %>% setNames(names(theta))
  
  .mod <- param(mod,.par)
  
  if(pred) {
    out <- mrgsim(.mod,data=.data,carry.out="type")
    return(as_data_frame(out))
  }
  
  out <- mrgsim(.mod, data=.data, obsonly=TRUE, Req="CP")
  
  return(wss(yobs,out$CP))
  
  #return(-1*sum(dnorm(log(yobs),log(out$CP),.par$sigma,log=TRUE)))
  
}
```


# Optimize with different methods

## `minqa::newuoa`: minimization without derivatives

- These appear to be the parameters that the authors are fitting

```{r}
theta <- log(c(fbCLintall = 1.2, ikiu = 1.2, 
               fbile = 0.9, ka = 0.1, ktr = 0.1))

control <- list(iprint=25)

fit1 <- newuoa(theta, pred,.data=data, yobs=yobs,control=control)

fit1$par <- setNames(fit1$par,names(theta))
```

### Get some predictions to look at how the fit went

- Predictions with the final estimates
- Predications with the initial estimates
- Observed data to overlay

```{r}
df_pred <- pred(fit1$par,dose,pred=TRUE) %>% mutate(type = typef(type))
df_init <- pred(theta,dose,pred=TRUE) %>% mutate(type = typef(type))
df_obs <- mutate(data, type=typef(type))
```

### Plot
```{r}
ggplot(data=df_pred) + 
  geom_line(data=df_init,aes(time,CP,lty="A"), col="black", lwd=0.7) +
  geom_line(aes(time,CP,lty="B"),col="darkslateblue",lwd=0.7) + 
  geom_point(data=df_obs,aes(time,DV),col="firebrick",size=2) + 
  facet_wrap(~type) + 
  scale_y_continuous(trans="log",breaks=10^seq(-4,4), 
                     limits=c(0.1,100),
                     "Pitavastatin concentration (ng/mL)") +
  scale_x_continuous(name="Time (hours)", breaks=seq(0,14,2)) +
  scale_linetype_manual(values= c(2,1),
                        labels=c("Initial estimates", "Final estimates"), name="") +
  theme(legend.position="top")  + theme_bw()
```

### The final objective function value and estimates
```{r}
pred(fit1$par,.data=data,yobs=yobs)

exp(fit1$par)
```

## `optim`: Nelder-Mead

```{r}
fit1b <- optim(theta, pred,.data=data, yobs=yobs)
```

## `DEoptim`: differential evolution algorithm

"Performs evolutionary global optimization via the 
Differential Evolution algorithm."

```{r, eval = TRUE}
lower <- rep(-6,length(theta)) %>% setNames(names(theta))
upper <- rep(4,length(theta)) %>% setNames(names(theta))

set.seed(330303)
decontrol <- DEoptim.control(NP=10*length(theta), CR=0.925, F=0.85,
                           itermax=90, storepopfrom=0)

fit2 <- DEoptim(fn=pred, lower=lower,upper=upper, control=decontrol,
                .data=data, yobs=yobs)

data.frame(initial = exp(theta),
           DE = exp(fit2$optim$bestmem),
           newuoa  = exp(fit1$par)) %>% signif(3)
```

### DA for the plot
```{r, eval = TRUE}
pops <- lapply(fit2$member$storepop, as.data.frame)
hx <- bind_rows(pops)
hx <- mutate(hx, iteration=rep(1:decontrol$itermax,each=decontrol$NP))
hx <- mutate(hx, pop = rep(1:decontrol$NP, time=decontrol$itermax))
hxm <- gather(hx, variable, value, 1:5) %>% mutate(value = exp(value))
best <- as_data_frame(fit2$member$bestmemit) %>% 
  mutate(iteration = 1:decontrol$itermax)
bestm <- gather(best,variable,value,1:5) %>% mutate(value = exp(value))
```

```{r, eval = TRUE}
ggplot(data=hxm) + 
  geom_line(aes(iteration,value,group=pop),col="darkslateblue") + 
  geom_line(data=bestm,aes(iteration,value),col="orange",lwd=1) + 
  scale_y_continuous(trans="log", breaks=10^seq(-4,4), name="Parameter value") + 
  facet_wrap(~variable, ncol=2, scales="free_y") 

```


## `GenSA`: simulated annealing 
```{r}
set.seed(11001)

sacontrol <- list(maxit = 100, nb.stop.improvement = 20, verbose = TRUE)

fit3 <- GenSA(NULL, pred, lower, upper, .data = data, yobs = yobs, 
              control = sacontrol)
```

## `hydroPSO`: particle swarm optimization
```{r}
set.seed(2202201)

fit4 <- hydroPSO(theta, fn = "pred", lower = lower, upper = upper, 
                 control = list(maxit = 100, REPORT = 5),
                 .data = data, yobs = yobs)
```

# Compare optimization methods
```{r}
results <- list(theta, fit1$par, fit1b$par, fit2$optim$bestmem, fit3$par, fit4$par)

results <- map(results, exp)

data_frame(
  method = c("initial", "newuoa", "nelder", "RcppDE", "SA", "PSO"),
  fbCLintall = map_dbl(results, "fbCLintall"), 
  ikiu = map_dbl(results, "ikiu"), 
  fbile = map_dbl(results, "fbile"), 
  ka = map_dbl(results, "ka"), 
  ktr = map_dbl(results, "ktr")
) %>% mutate_if(is.numeric,signif,digits = 4) %>% knitr::kable()

```

