---
title: A multiscale systems model of bone health and mineral homeostasis
author: "Metrum Research Group, LLC"
date: ""
output: 
  github_document:
    toc: true
---

```{r, echo = FALSE}
knitr::opts_chunk$set(comment = '.', warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.path = "img/OpenBoneMin-", fig.width = 8)
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.path = "cache/bone/")

```

# Background and Motivation

The model was originally developed to describe the bone marker changes associated with denosumab administration from a then ongoing clinical trial. Associated changes in serum calcium and PTH were also considered of interest at the time and so justified the development of a 'systems' model that included bone remodeling and bone mineral (calcium and phosphate) homeostatic mechanisms. Other therapeutics (e.g., teriparatide) and disease states (kidney failure, parathyroid-related abnormalities) were also considered at the time to further inform the parameterization and estimation of the model (Peterson and Riggs, Bone 2010).

# Install from GitHub

```{r,eval = FALSE}
remotes::install_github("metrumresearchgroup/OpenBoneMin")
```

```{r}
library(OpenBoneMin)
library(tidyverse)
```


# Load the Bone / Mineral model
```{r}
mod <- BoneMin()
```


# Teriparatide example

- Teriparatide is a recombinant parathyroid (PTH) hormone, containing only the 
first 84 amino acid residues
- Teriparatide doses are administered subcutaneously and are absorbed into
the PTH compartment; so teriparatide is considered equivalent to PTH 
- We'll give either 20 or 40 micrograms SQ daily for 10 doses
```{r}
out <- sim_teri(dose = c(20,40), dur = 9)

out
```

PTH profiles for the 20 and 40 microgram doses

```{r}
plot(out, PTHpm ~ time|ID, scales = "same")
```


Calcium profiles for the 20 and 40 microgram doses

```{r}
plot(out, CaC~time)
```


# Denosumab example

- Denosumab is a human monoclonal antibody that binds RANK ligand (RANKL), 
preventing the formation of RANK/RANKL complex and subsequent simulation 
of osteoclast maturation
- The labeled dose for treatment of osteoporosis is 60 mg SQ every 6 months

```{r}
out <- sim_denos(dose = c(14,30,60,210))
```

Plot the denosumab concentration versus time

```{r}
plot(out, DENCP ~.,  scales = list(y=list(log = TRUE)), ylim = c(1E-4, 10E5))
```

Plot the percent change in lumbar spine BMD versus time
```{r}
plot(out, BMDlsDENchange ~ .)
```

# Peterson and Riggs 2012


__Predicting Nonlinear Changes in Bone Mineral Density Over Time Using a 
Multiscale Systems Pharmacology Model__

Peterson MC, Riggs MM. CPT Pharmacometrics Syst Pharmacol. 2012 Nov 14;1:e14. 
doi: 10.1038/psp.2012.15. PubMed PMID: 23835796; PubMed Central PMCID: 
PMC3600731 [PubMed Citation](https://www.ncbi.nlm.nih.gov/pubmed/23835796) / [Free 
Full Text](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3600731)


- Bone is constantly being remodeled, constantly turning over.  This remodeling 
  process consists of breaking down old bone and building up new bone
    - *Osteoclasts* are cells that function to disassemble bone (resorption) and 
      *osteoblasts* are cells that function to synthesize or build new bone
    - One therapeutic strategy for osteoporosis (porous bone) is to inhibit 
      osteoclast activity and thus inhibit the bone resorption  activity  
    - __Denosumab__ decreases osteoclast activity by preventing maturation of
      pre-osteoclasts into fully-functional osteoclast cells

- The goal of the modeling was to predict nonlinear changes in lumbar spine (LS)
  bone mineral density (BMD) over 48 months of denosumab treatment under
  different dosing regimens. 

- __LSBMD__ was modeled as a function of bone turnover markers
     - CTx: C-telopeptide, a marker for osteoclast function
     - BSAP: bone-specific alkaline phosphatase, a marker for osteoblast 
       function

- Our __simulation objective__ is to recreate figures 3, 4, 5, and 6 from 
the publication.  This will show the changes in CTx and BSAP and the 
subsequent changes in BMD under the model.   

## Generate the dosing regimens

__60 mg every 6 months x 8__

```{r}
month <- 24*28
regi <- function(x) factor(x,labels = c("60 mg Q6M", "14 mg Q6M", 
                                        "30 mg Q3M", "210 mg Q6M"))
xscale <- scale_x_continuous(breaks = c(0,3,6,12,18,24,36,48), limits = c(0,48))
```

```{r}

e1 <- ev(amt = 60, ii = 6*month, addl = 7)
e1
```

__14 mg every 6 months x4, then 60 mg every 6 months x4__
```{r}
e2 <- 
  mutate(e1, amt = 14, addl = 3) %then% 
  mutate(e1, addl = 3)
e2
```

__30 mg every 3 months for 8 doses and changed to 60 mg Q6M starting 
on month 36__

```{r}
e3a <- ev(amt = 30, ii = 3*month, addl = 7)
e3b <- mutate(e1, addl = 1)
e3 <- seq(e3a, wait = 12*month, e3b)
e3
```


__210 mg every 6 months x4 then DC__
```{r}
e4 <- mutate(e1, amt = 210, addl = 3)
e4
```


__Create one single data set from which to simulate__
```{r}
data <- as_data_set(e1,e2,e3,e4)

data
```

```{r}
out <- 
  mod %>%
  mrgsim(data = data, end = 48*month, delta = 24) %>% 
  mutate(ID = regi(ID), Month = time/month)
```

## Publication figure 3: CTx vs time

Shown here is the percent change from baseline value for osteoclasts (via CTx)

```{r}
ggplot(out) + 
  geom_line(aes(x = Month, y = OCchange, col = factor(ID)), lwd = 1) + 
  facet_grid(~ID) + geom_hline(yintercept = 100, lty = 2) +
  scale_y_continuous(trans = "log10", breaks = c(10,30,100,300), limits = c(5,300)) + 
  theme_bw() + theme(legend.position = "top") +
   geom_vline(xintercept = c(24,36), lty = 3) + xscale
```

```{r, echo = FALSE}
p <- ggplot(out) + 
  geom_line(aes(x = Month, y = OCchange, col = factor(ID)), lwd = 1) + 
  facet_wrap(~ID, ncol = 2) + geom_hline(yintercept = 100, lty = 2) +
  scale_y_continuous(trans = "log10", breaks = c(10,30,100,300), limits = c(5,300)) + 
  theme_bw() + theme(legend.position = "none") +
   geom_vline(xintercept = c(24,36), lty = 3) + xscale

p <- p + labs(y = "Osteoblasts", color = "", subtitle = "Denosumab for osteoporosis")
saveRDS(p, file = "img/README_osteo.RDS")
```


## Publication figure 4: BSAP vs time

Shown here is the percent change from baseline value for osteoblasts (via BSAP)

```{r}
ggplot(out) + 
  geom_line(aes(x = Month, y = OBchange, col = factor(ID)), lwd = 1) + 
  facet_grid(~ID) + geom_hline(yintercept = 100, lty = 2) +
  scale_y_continuous(trans = "log10", breaks = c(10,30,100,300), limits = c(5,300)) +
  theme_bw() + theme(legend.position = "top") + 
  geom_vline(xintercept = c(24,36), lty = 3) + xscale
```

## Publication figure 5: LS BMD vs time

Shown here is the percent change from baseline value for lumbar spine BMD

```{r}
ggplot(out) + 
  geom_line(aes(x = Month, y = BMDlsDENchange, col = factor(ID)), lwd = 1) + 
  facet_grid(~ID) + 
  theme_bw() + theme(legend.position = "top") + 
  geom_vline(xintercept = c(24,36), lty = 3) + xscale
``` 


## Publication figure 6: TGF-$\beta$ vs time

- Solid lines are __active__ TGF-$\beta$
- Dashed lines are __latent__ TGF-$\beta$

```{r}
out <- 
  out %>%
  group_by(ID) %>%
  mutate(ATGF = 100*TGFBact/first(TGFBact), LTGF = 100*TGFB/first(TGFB)) %>%
  ungroup

ggplot(out) + 
  geom_line(aes(x = Month, y = ATGF, col = factor(ID)), lwd = 1) + 
  geom_line(aes(x = Month, y = LTGF), col = "black", lty = 2, lwd = 0.6) + 
  facet_grid(~ID) + 
  theme_bw() + theme(legend.position = "top") + 
  geom_vline(xintercept = c(24,36), lty = 3) + xscale +
  scale_y_continuous(breaks = c(0,25,50,75,100,150,225), limits = c(0,225)) 
```