---
title: A multiscale systems model of bone health and mineral homeostasis
author: "Metrum Research Group, LLC"
date: ""
output: 
  github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(comment = '.', warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.path = "img/OpenBoneMin-")
```

# Background and Motivation

The model was originally developed to describe the bone marker changes associated with denosumab administration from a then ongoing clinical trial. Associated changes in serum calcium and PTH were also considered of interest at the time and so justified the development of a 'systems' model that included bone remodeling and bone mineral (calcium and phosphate) homeostatic mechanisms. Other therapeutics (e.g., teriparatide) and disease states (kidney failure, parathyroid-related abnormalities) were also considered at the time to further inform the parameterization and estimation of the model (Peterson and Riggs, Bone 2010).

# Install from GitHub

```{r,eval = FALSE}
remotes::install_github("metrumresearchgroup/OpenBoneMin")
```

```{r}
library(OpenBoneMin)
library(dplyr)
```


# Load the Bone / Mineral model
```{r}
mod <- BoneMin()
```


# An example simulation of teriparatide administration

- We'll give either 20 or 40 micrograms SQ daily for 
```{r}
out <- sim_teri(dose = c(20,40), dur = 9)

out
```

PTH profiles for the 20 and 40 microgram doses

```{r}
plot(out, PTHpm ~ time|ID, scales = "same")
```


Calcium profiles for the 20 and 40 microgram doses

```{r}
plot(out, CaC~time)
```


# Denosumab administration
```{r}
out <- sim_denos(dose = c(30,60,210))

plot(out, DENCP ~.,  scales = list(y=list(log = TRUE)), 
     ylim = c(1E-4, 10E5))
```
```{r}
plot(out, BMDlsDENchange ~ .)
```

