---
title: "Rifampicin PBPK Model to Predict Complex DDIs"
author: "Metrum Research Group, LLC"
output: 
  github_document: 
   toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '.', warning = FALSE, message = FALSE, 
                      fig.path = "img/rifampin_midazolam_ddi-")
knitr::opts_chunk$set(cache=TRUE, autodep=TRUE, cache.path = "cache/rifampin/")
```

```{r}
library(mrgsolve)
library(PKPDmisc)
theme_set(theme_bw())
theme_update(legend.position = "top")
```

# Reference
This model and vignette was derived from this publication

__Comprehensive PBPK Model of Rifampicin for Quantitative Prediction of
Complex Drug-Drug Interactions: CYP3A/2C9 Induction and OATP Inhibition Effects__
    - Asaumi R, Toshimoto K, Tobe Y, Hashizume K, Nunoya KI, Imawaka H, Lee W,
      Sugiyama Y.  CPT Pharmacometrics Syst Pharmacol. 2018 Jan 25. 
      PMID: 29368402
      - https://www.ncbi.nlm.nih.gov/pubmed/29368402



# Rifampin PBPK

```{r}
mod <- mread_cache("rifampicin_midazolam", "models", delta = 0.1)
```


## Single rifampicin dose
```{r}
rif <- ev(amt = 600)
rif
```


```{r}
mod %>%
  ev(rif) %>% 
  Req(Ccentral) %>%
  mrgsim(end = 48) %>% 
  plot()
```


## Multiple rifampicin doses
```{r}
rif <- mutate(rif, ii = 24, addl = 9)

rif

out <- 
  mod %>%
  ev(rif) %>% 
  mrgsim(end = 240)
```

What is going on here?
```{r}
plot(out, Ccentral ~ time)
```


Let's investigate

```{r}
aucs <- 
  out %>% 
  mutate(DAY = 1+floor(time/24)) %>%
  group_by(DAY) %>% 
  summarise(AUC = auc_partial(time,Ccentral)) %>% 
  ungroup %>% 
  mutate(pAUC = 100*AUC/first(AUC)) %>%
  filter(DAY < 10)

ggplot(aucs, aes(factor(DAY),pAUC)) + 
  geom_col(alpha = 0.6) + 
  geom_hline(yintercept = 70, lty = 2, col = "firebrick") + 
  scale_y_continuous(breaks = seq(0,100,10)) + 
  ggtitle("Auto-induction of rifampicin metabolism")
```


Both `CYP3A4` and `UGT` metabolic activity increased after multiple
rifampicin doses

```{r}
simsm <- 
  out %>%
  as_data_frame() %>%
  gather(variable, value, c(UGT_ratio_HC5, CYP3A4_ratio_HC5))

simsm %>%
  ggplot(., aes(time/24, value, col = variable)) + 
  geom_line(lwd =1 ) + scale_color_brewer(palette = "Set2") 
```

# PBPK model for rifampicin / midazolam DDI

- Recall that our PBPK model is really a combination of two models:
one for rifampicin and one for midazolam


A single 3 mg midazolam dose

```{r}
mid <- ev(amt = 3, cmt = 2)

mid

mod %>% mrgsim_e(mid, Req = "Cmidazolam") %>% plot()
```

Now, a single 3 mg midazolam dose after 7 days of rifampin 75 mg QD

```{r}
rif <- ev(amt = 75, ii = 24, addl = 6)

rif_mid <- ev_seq(rif, wait = -12, mid)

mid <- filter(rif_mid, cmt==2)

both <- as_data_set(mid, rif_mid)

both
```

```{r}
sims <- 
  mod %>% 
  mrgsim_d(both, Req="Cmidazolam", end = 166) %>% 
  filter_sims(time >= 156) %>% 
  mutate(ID = factor(ID, labels = c("Midazolam", "Midazolam after Rif")))


ggplot(sims, aes(time,Cmidazolam,col = factor(ID))) + 
  geom_line(lwd = 1) + 
  scale_y_continuous(trans = "log10", limits = c(0.1, 10)) +
  scale_color_brewer(palette = "Set2")
```

Midazolam exposure is reduced after rifampicin 75 mg daily x 7d

```{r}
sims %>% 
  group_by(ID) %>% 
  summarise(AUC = auc_partial(time,Cmidazolam)) %>%
  mutate(percent_reduction = 100*(1-AUC/first(AUC))) %>%
  knitr::kable(digits = 3)
```

## Dose-response for midazolam/rifampin DDI

Make a function to wrap up the workflow for a single dose

```{r}

sim_ddi <- function(rif_dose, mid_dose = 3) {
  mid <- ev(amt = mid_dose, cmt = 2)
  rif <- ev(amt = rif_dose, ii = 24, addl = 6)
  rif_mid <- ev_seq(rif, wait = -12, mid)
  
  mod %>% 
    mrgsim_e(rif_mid, Req="Cmidazolam", end = 166) %>% 
    filter_sims(time >= 156) %>%
    mutate(rif = rif_dose, mid = mid_dose)
}


out <- map_df(seq(0,600,10), .f = sim_ddi)
```

Summarize the simulations by both rifampicin dose and midazolam dose.  Because
we simulated the zero rifampicin dose, we get the percent reduction in AUC
by dividing by the "first" auc in the series

```{r}
summ <- 
  out %>%
  group_by(rif,mid) %>%
  summarise(auc = auc_partial(time,Cmidazolam)) %>%
  ungroup() %>%
  mutate(pAUC = 100*(1-auc/first(auc)))

summ
```


```{r}
ggplot(summ, aes(rif,pAUC)) + 
  geom_line(lwd = 1) + 
  scale_y_continuous(breaks = seq(0,100,10), limits = c(0,100),
                     name = "%Reduction in midazolam AUC after Rif x 7d") + 
  scale_x_continuous(name = "Rifampicin dose (mg)", 
                     breaks = seq(0,600,100)) + theme_bw()
```





